---
title: "Exercise 2.6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Remark:** Before running this code in RStudio, first do (from the menu): *Session > Set Working Directory > To Source File Location*.
To use the glimpse function, you need to install the package dplyr (*Tools > Install Packages*).

```{r}
library(dplyr)
```

Import the file `android_persistence_cpu.csv` in R.


```{r}
android_cpu <- read.csv("../../datasets/android_persistence_cpu.csv", sep = ";")
attach(android_cpu)
```
    
1. Get an overview of the different variables in this dataset


    ```{r}
    ncol(android_cpu)       # Number of variables
    colnames(android_cpu)   # Variable names
    ```
    
2. Use the functions `glimpse` and `View` to view the contents of the dataset.

    ```{r}
    #View(android_cpu)  
    glimpse(android_cpu)
    ```

3. Use the function `unique` to get a list of distinct values for each variable in the sample.

    ```{r}
    unique(android_cpu$Time)
    unique(android_cpu$PersistenceType)
    unique(android_cpu$DataSize)
    ```
    
4. Define the level of measurement for each variable. The results from the previous steps can help!

    - `Time`: quantative **ratio**, there is a non-arbitrary zero point
    - `PersistenceType`: qualitative **nominal**, there is no logical order between the values
    - `DataSize`: qualitative: **ordinal**, can be ordered (e.g. Small - Medium - Large)
