---
title: "Exercise 2.6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Remark:** Before running this code in RStudio, first do (from the menu): *Session > Set Working Directory > To Source File Location*.


Import the file `earthquakes.csv` in R.


```{r}
earthquakes <- read.csv("../../datasets/earthquakes.csv", sep = ",")
attach(earthquakes)
```
    
What is the measurement level of the following variables?

1. Latitude and Longitude

    ```{r}
    range(earthquakes$Latitude)          # Range
    length(unique(earthquakes$Latitude)) # Count number of unique values
    
    range(earthquakes$Longitude)
    length(unique(earthquakes$Longitude))
    ```
    
    Both variables use an **interval quantitative** scale.
    
2. Type

    ```{r}
    unique(earthquakes$Type)
    ```
    
    This is a **nominal qualitative** variable.
    
3. Time

    ```{r}
    length(unique(earthquakes$Time))
    ```
    
    Time is an **interval quantitative** scale.
    
4. Depth

    ```{r}
    range(earthquakes$Depth)
    length(unique(earthquakes$Depth))
    ```
    
    This variable uses a **ratio quantitative** scale.
    
