---
title: "Exercise 3.13-19."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Remark:** Before running this code in RStudio, first check the following setting (from the menu): *Session > Set Working Directory > To Source File Location*. This allows for loading the .csv file from the current directory.

# Case Study: Android Persistence CPU

## Loading the data

```{r}
# Load data from csv
android_cpu <- read.csv("..datasets/android_persistence_cpu.csv", sep = ";")
# Allows the columns to be accessed directly instead of with table$column notation
attach(android_cpu)
```

## Descriptive statistics

```{r}
summary(android_cpu)
```

What are the possible values for `DataSize` and `PersistenceType`?

```{r}
levels(DataSize)
levels(PersistenceType)
```

Frequency table for `DataSize`, `PersistenceType`

```{r}
table(DataSize)
table(PersistenceType)
table(DataSize, PersistenceType)
```

Mean, variance, standard deviation over the entire data set

```{r}
mean(Time)
var(Time)
sd(Time)
```

The same, but over the different persistence types

```{r}
aggregate(Time ~ PersistenceType, FUN = mean)
aggregate(Time ~ PersistenceType, FUN = var)
aggregate(Time ~ PersistenceType, FUN = sd)
```

The same, but over the data size

```{r}
aggregate(Time ~ DataSize, FUN = mean)
aggregate(Time ~ DataSize, FUN = var)
aggregate(Time ~ DataSize, FUN = sd)
aggregate(Time ~ PersistenceType * DataSize, 
          FUN = mean)
```

Median and quantiles

```{r}
aggregate(Time ~ PersistenceType, FUN = quantile)
aggregate(Time ~ DataSize, FUN = quantile)
```

## Data visualisation

### Histogram

cfr. Exercise 3.14

```{r}
hist(Time, main="Distribution of execution time", 
     xlab="Execution time (ms)");
hist(Time, main="Distribution of execution time",
     xlab="Execution time (ms)", breaks=2);
```

### Box Plot

cfr. Exercise 3.15 

```{r}

boxplot(Time)                   # Or boxplot(android_cpu$Time)
boxplot(Time,
        main='Distribution of execution time',
        ylab="Execution time (ms)");
```

We can set `horizontal` to `TRUE` to draw a horizontal boxplot. 
Note that in this case we need to add the label to the `x` axis.

```{r}

boxplot(Time,
        horizontal = TRUE,
        main='Distribution of execution time',
        xlab="Execution time (ms)");
```

cfr. Exercise 3.16
        
```{r}


boxplot(Time ~ DataSize)
boxplot(Time ~ DataSize,
        main="Distribution of CPU time over the data sizes",
        ylab="Time in ms")
```

cfr. Exercise 3.17
```{r}
boxplot(Time ~ PersistenceType * DataSize,
        main="Distribution of CPU time over data sizes for the different persistent types",
        ylab="Time in ms")
```

Data selection using which
cfr. Exercise 3.18 - 3.19

```{r}
greenDOA <- android_cpu[which(android_cpu$PersistenceType=='GreenDAO'),];
boxplot(greenDOA$Time ~ greenDOA$DataSize,
        main="Distribution of execution time for GreenDAO");
```

Data selection using subset

```{r}
subset_small = subset(android_cpu, DataSize == 'Small')
boxplot(subset_small$Time ~ subset_small$PersistenceType,
        main = "Execution time of the small experiments")

subset_medium = subset(android_cpu, DataSize == 'Medium',
        main = "Execution time of the medium size experiments")
boxplot(subset_medium$Time ~ subset_medium$PersistenceType)

subset_large = subset(android_cpu, DataSize == 'Large',
        main = "Execution time of the large experiments")
boxplot(subset_large$Time ~ subset_large$PersistenceType)
```

Data selection using which (same results)

```{r}
results_small <- android_cpu[which(android_cpu$DataSize=='Small'),]
boxplot(results_small$Time ~ results_small$PersistenceType,
        main = "Execution time of the small experiments")

results_medium <- android_cpu[which(android_cpu$DataSize=='Medium'),]
boxplot(results_medium$Time ~ results_medium$PersistenceType,
        main = "Execution time of the medium size experiments")

results_large <- android_cpu[which(android_cpu$DataSize=='Large'),]
boxplot(results_large$Time ~ results_large$PersistenceType,
        main = "Execution time of the large experiments")
```
