---
title: "3 -- Univariate Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quantitative variables

Some basic functions for univariate analysis on a quantitative variable.

```{r}
# Lengths superheroes
lengths <- c(141, 198, 143, 201, 184)
```

### Measures of central tendency

```{r}
mean(lengths)                     # mean or average
median(lengths)
```

### Measures of dispersion

```{r}
range(lengths)                    # minimum & maximum
abs(max(lengths) - min(lengths))  # range
summary(lengths)                  # Quartiles, etc.
IQR(lengths)                      # Interquartile range
sd(lengths)                       # standard deviation of a *sample*
```

### Formula breakdown

Mathematical formulae can often be translated to R quite straightforwardly. Take e.g. the formula for the sample mean:

$\overline{x} = \frac{\sum_{x=1}^{n}x_i}{n}$

In R, this becomes:

```{r}
sum(lengths) / length(lengths)
```

The same goes for the variance and standard deviation of a sample. The sample variance is given by:

$s^2 = \frac{\sum_{i=1}^{n-1} \left(x_{i} - \overline{x}\right)^2}{n-1}$

In the code below, we break down the formula in parts:

```{r}
res_1 <- lengths - mean(lengths)    # difference of data points with mean
res_2 <- res_1^2                    # squared differences
res_3 <- sum(res_2)                 # take the sum
variance <- res_3 / length(lengths) # calculate average
stdev <- sqrt(variance)             # take the square root
```

Or, the entire calculation of variance in one formula. Can you recognize the mathematical formula in the R-code?

```{r}
variance <- sum((lengths - mean(lengths))^2) / (length(lengths) - 1)
```

### Calculations by group

Often, you want to get calculations of mean, standard deviation, etc. grouped by some factor. This can be done using the `aggregate` function, and the "group by" `~` operator.

We show some examples from the `mtcars` dataset.

```{r}
View(mtcars)
# Show the average mileage per gallon for cars with and
# without an automatic transmission.
aggregate(mpg ~ am, data = mtcars, FUN = mean)
# Standard deviation
aggregate(mpg ~ am, data = mtcars, FUN = sd)
# Apply the summary function
aggregate(mpg ~ am, data = mtcars, FUN = summary)
```

Remark that the `data` parameter allows you to reference the column names directly, instead of having to use the notation `mtcars$mpg ~ mtcars$am`.

## Qualitative variables

In R, qualitative variables are called *factors*. As an example, we'll use the `esoph` dataset from the `datasets` package, available in R.

```{r}
?esoph
View(esoph)
```

### Measures of central tendency

About the only measure of central tendency for a factor/qualitative variable is the mode. There is no actual mode function in R, but you can find it in several ways. The first is to print a frequency table and read the maximum from there:

```{r}
levels(esoph$agegp)            # Show all values of the variable
freq_tab <- table(esoph$agegp) # Calculate a frequency table
freq_tab
summary(esoph$agegp)           # The summary function applied to a factor
```

The output of both `table` and `summary` (applied to a factor) is identical.

From there, we could look for the value that occurs the most:

```{r}
which.max(table(esoph$agegp)) # Only works for the case with a single mode
names(freq_tab)[freq_tab == max(freq_tab)] # Also works for multimodal variables
```

It's a bit convoluted, but in R, this is the only way to calculate the mode.


