---
title: "3 -- Univariate Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quantitative variables

Some basic functions for univariate analysis on a quantitative variable.

```{r}
# Lengths superheroes
lengths <- c(141, 198, 143, 201, 184)
```

### Measures of central tendency

```{r}
mean(lengths)                     # mean or average
median(lengths)
```

### Measures of dispersion

```{r}
range(lengths)                    # minimum & maximum
abs(max(lengths) - min(lengths))  # range
summary(lengths)                  # Quartiles, etc.
IQR(lengths)                      # Interquartile range
sd(lengths)                       # standard deviation of a *sample*
```

### Formula breakdown

Mathematical formulae can often be translated to R quite straightforwardly. Take e.g. the formula for the sample mean:

$\overline{x} = \frac{\sum_{x=1}^{n}x_i}{n}$

In R, this becomes:

```{r}
sum(lengths) / length(lengths)
```

The same goes for the variance and standard deviation of a sample. The sample variance is given by:

$s^2 = \frac{\sum_{i=1}^{n-1} \left(x_{i} - \overline{x}\right)^2}{n-1}$

In the code below, we break down the formula in parts:

```{r}
res_1 <- lengths - mean(lengths)    # difference of data points with mean
res_2 <- res_1^2                    # squared differences
res_3 <- sum(res_2)                 # take the sum
variance <- res_3 / (length(lengths)-1) # calculate average
stdev <- sqrt(variance)             # take the square root
```

Or, the entire calculation of variance in one formula. Can you recognize the mathematical formula in the R-code?

```{r}
variance <- sum((lengths - mean(lengths))^2) / (length(lengths) - 1)
```

### Calculations by group

Often, you want to get calculations of mean, standard deviation, etc. grouped by some factor. This can be done using the `aggregate` function, and the "group by" `~` operator.

We show some examples from the `mtcars` dataset.

```{r}
View(mtcars)
# Show the average mileage per gallon for cars with and
# without an automatic transmission.
aggregate(mpg ~ am, data = mtcars, FUN = mean)
# Standard deviation
aggregate(mpg ~ am, data = mtcars, FUN = sd)
# Apply the summary function
aggregate(mpg ~ am, data = mtcars, FUN = summary)
```

Remark that the `data` parameter allows you to reference the column names directly, instead of having to use the notation `mtcars$mpg ~ mtcars$am`.

## Qualitative variables

In R, qualitative variables are called *factors*. As an example, we'll use the `esoph` dataset from the `datasets` package, available in R.

```{r}
?esoph
View(esoph)
```

### Measures of central tendency

About the only measure of central tendency for a factor/qualitative variable is the mode. There is no actual mode function in R, but you can find it in several ways. The first is to print a frequency table and read the maximum from there:

```{r}
levels(esoph$agegp)            # Show all values of the variable
unique(esoph$agegp)
freq_tab <- table(esoph$agegp) # Calculate a frequency table
freq_tab
summary(esoph$agegp)           # The summary function applied to a factor
```

The output of both `table` and `summary` (applied to a factor) is identical.

From there, we could look for the value that occurs the most:

```{r}
which.max(table(esoph$agegp)) # Only works for the case with a single mode
names(freq_tab)[freq_tab == max(freq_tab)] # Also works for multimodal variables
```

It's a bit convoluted, but in R, this is the only way to calculate the mode.

## Sample variance vs population variance

As stated above, the sample variance is given by $s^2 = \frac{\sum_{i=1}^{n-1} \left(x_{i} - \overline{x}\right)^2}{n-1}$. The $n-1$ in the denominator may seem unintuitive: why not $n$? Indeed, the formula with $n$ as denominator does exist and is known as the population variance. Why do we use a different formula for the sample variance? It can be proven mathematically that in a sample, using $n-1$ in the denominator gives an unbiased estimation of the (unknown) population variance. The actual proof is outside of the scope of this course, but we'll illustrate it with an example. We'll use both formulas to calculate the variance of a sample and will compare it with the actual population variance. The formula that, on average, provides the best estimate is the winner!

First, let's define functions for population variance and standard deviation. These are not supported out-of-the-box in R. The following functions calculate the population variance and standard deviation, respectively:

```{r}
# Population variance
pop_var <- function(x) {
  mean <- sum(x) / length(x)
  1/length(x) * sum ((x - mean)^2)
}

# Population standard deviation
pop_sd <- function(x) {
  sqrt(pop_var(x))
}
```

Let's say our population consists of the numbers 0 - 100

```{r}
population <- 0:100
```

First, we calculate the mean and standard deviation of the population

```{r}
# µ = 50
population_mean <- sum(population) / length(population)
# σ² = 850
population_variance <- pop_var(population)
# σ ≈ 29.15
population_stdev <- pop_sd(population)

sprintf("Population mean    : %f", population_mean)
sprintf("Population variance: %f", population_variance)
sprintf("Population stdev   : %f", population_stdev)
```

We will show that the formula with denominator $n-1$ gives a better estimate of the actual population variance than denominator $n$.

We will take random samples from the population, calculate the variance with both formulas, and see which one is closest to the actual value.

```{r}
sample_size <- 15
num_samples <- 5000

pop_sample <- function(x) sample(population, sample_size)
samples <- lapply(1:num_samples, pop_sample)
```

First, let's calculate standard deviations with the "population" formula: denominator $n$:

```{r}
results_pop_sd <- sapply(samples, pop_sd)
average_n <- mean(results_pop_sd)
sprintf("denominator n  : %f, difference : %f", 
        average_n, abs(average_n - population_stdev))
```

Then, calculate standard deviation with the built-in function sd(): denominator $n-1$:

```{r}
results_sample_sd <- sapply(samples, sd)
average_n_min_1 <- mean(results_sample_sd)

sprintf("denominator n-1: %f, difference : %f", 
        average_n_min_1, abs(average_n - average_n_min_1))
```

On average, the formula with denominator $n-1$ gives a better estimate!
