---
title: "3 -- Univariate Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quantitative variables

Some basic functions for univariate analysis on a quantitative variable.

```{r}
# Lengths superheroes
lengths <- c(141, 198, 143, 201, 184)
```

### Measures of central tendency

```{r}
mean(lengths)                     # mean or average
median(lengths)
```

### Measures of dispersion

```{r}
range(lengths)                    # minimum & maximum
abs(max(lengths) - min(lengths))  # range
summary(lengths)                  # Quartiles, etc.
IQR(lengths)                      # Interquartile range
sd(lengths)                       # standard deviation of a *sample*
```

### Formula breakdown

Mathematical formulae can often be translated to R quite straightforwardly. Take e.g. the formula for the mean:

$\mu = \frac{\sum_{x=1}^{n}x_i}{n}$

In R, this becomes:

```{r}
sum(lengths) / length(lengths)
```

The same goes for the variance and standard deviation. In the example below, we use the definition of population variance (denominator $n$).

$\sigma^2 = \frac{\sum_{i=1}^{n} \left(x_{i} - \mu\right)^2}{n}$

In the code below, we break down the formula in parts:

```{r}
res_1 <- lengths - mean(lengths)    # difference of data points with mean
res_2 <- res_1^2                    # squared differences
res_3 <- sum(res_2)                 # take the sum
variance <- res_3 / length(lengths) # calculate average
stdev <- sqrt(variance)             # take the square root
```

Or, the entire calculation of variance in one formula. Can you recognize the mathematical formula in the R-code?

```{r}
variance <- sum((lengths - mean(lengths))^2) / length(lengths)
```

### Calculations by group

Often, you want to get calculations of mean, standard deviation, etc. grouped by some factor. This can be done using the `aggregate` function, and the "group by" `~` operator.

We show some examples from the `mtcars` dataset.

```{r}
View(mtcars)
# Show the average mileage per gallon for cars with and
# without an automatic transmission.
aggregate(mpg ~ am, data = mtcars, FUN = mean)
# Standard deviation
aggregate(mpg ~ am, data = mtcars, FUN = sd)
# Apply the summary function
aggregate(mpg ~ am, data = mtcars, FUN = summary)
```

Remark that the `data` parameter allows you to reference the column names directly, instead of having to use the notation `mtcars$mpg ~ mtcars$am`.

## Qualitative variables

In R, qualitative variables are called *factors*. As an example, we'll use the `esoph` dataset from the `datasets` package, available in R.

```{r}
?esoph
View(esoph)
```

### Measures of central tendency

About the only measure of central tendency for a factor/qualitative variable is the mode. There is no actual mode function in R, but you can find it in several ways. The first is to print a frequency table and read the maximum from there:

```{r}
levels(esoph$agegp)            # Show all values of the variable
freq_tab <- table(esoph$agegp) # Calculate a frequency table
freq_tab
summary(esoph$agegp)           # The summary function applied to a factor
```

The output of both `table` and `summary` (applied to a factor) is identical.

From there, we could look for the value that occurs the most:

```{r}
which.max(table(esoph$agegp)) # Only works for the case with a single mode
names(freq_tab)[freq_tab == max(freq_tab)] # Also works for multimodal variables
```

It's a bit convoluted, but in R, this is the only way to calculate the mode.

## Charts in R

For univariate statistics, the most common chart types are:

- **boxplot**, that shows the spread of a (quantitative)
- **bar chart**, which can be used for
    - showing the values of a quantitative variable
    - showing the frequencies of a qualitative variable
- **histogram**, a variant of the bar chart for frequencies, where ranges of x-values are taken together in "buckets"

```{r}
# Chart types
barplot(lengths)  # bar chart (without any fancy layout add-ons)
boxplot(lengths)  # boxplot

# Example "Active Duty Personnel, 1998"
active_duty_personnel <- c(492, 363, 381, 176)

barplot(active_duty_personnel)
pie(active_duty_personnel)       # Shown for reference, avoid pie charts!

# Plot of the frequencies of a qualitative variable
barplot(table(esoph$agegp))
```

As an example of a histogram, we take the `cars` dataset as an example. 

```{r}
# For reference, a boxplot of the data
boxplot(cars$dist, horizontal = TRUE)

# A simple histogram, the hist function decides on the number
# of buckets
hist(cars$dist)

# Only 4 "breaks" between buckets => 5 buckets
hist(cars$dist, breaks = 4)  

# Specify boundaries between buckets explicitly
hist(cars$dist, breaks = c(0,30,60,90,120))  
```


### More elaborate charts

Charts can be extended with titles, legends, colours, etc. A few examples are given here, you can find a lot more on the Internet. Remark that there is a separate chart library in R, called `ggplot`, that is not discussed in this guide.

```{r}
heroes <- c("Spiderman", "Batman", "Superman",
            "Deadpool", "Catwoman")
barplot(lengths,
        main = "Lengths of super heroes",
        names = heroes,
        xlab = "Hero",
        ylab = "Length (cm)",
        col = "light blue"
        )

military_categories <- c("Army", "Air Force", "Navy", "Marine Corps")
barplot(active_duty_personnel,
        names = military_categories)
```
